The Right Way to Search Evolving Graphs

### 2016-03-04 paper.tex
Describe Julia machine better
various minor fixes

It is well known that sparse matrix-vector product is equivalent to a
one-step BFS on the corresponding (static) graph~\cite[Sec. 1.1]{kegi11}.
In this paper, we demonstrate a correct corresponding result for an evolving
graph by constructing a block triangular matrix representation of the graph that
takes into account both static and causal connections between active nodes.
In Section~\ref{sec:breadth-first-search}, we explain how the BFS algorithm can
be applied to evolving graph to enumerate paths that traverse edges across time
and space. Section~\ref{sec:temporal-paths} provides an example showing that considering
only products of the time-dependent adjacency matrices fails to enumerate certain
temporal paths. We present and demonstrate the BFS algorithm over evolving graphs
in Section~\ref{sec:evolving-graph-bfs}, showing its formal equivalence to
BFS over a particular static graph generated by adding causal edges that connect
active nodes. This static graph generates an algebraic
representation of the BFS as power iteration of its adjacency matrix to a starting
search node, as shown in Section~\ref{sec:bfsla}. The algebraic formulation also
demonstrates interesting connections between properties of the BFS algorithm and
the adjacency matrix. We describe in Section~\ref{sec:implementation-julia} an
implementation of the algorithm in Julia. Finally in Section~\ref{sec:applications},
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

The node set $V$ and edge set $E$ now define a directed static graph $G=(V,E)$
that is in 1-1 correspondence with the evolving graph $G_n$. The node set $V$ of $G$ is in 1-1 correspondence with active nodes of $G_n$ while the edge set $E$ is in 1-1
correspondence with all temporal paths of length 2 on $G_n$.

@@

We can then represent $G_n$ using the sequence of adjacency matrices
$A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$. The example in
Figure~\ref{fig:eg_shortest_path} can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle.
\]

@@

We generate a sequence of random (directed) \texttt{IntEvolvingGraph}s
with $10^5$ active nodes and $10$ time stamps.
The first \texttt{IntEvolvingGraph} in the sequence has about
$10^8$ static edges.
We consecutively add new
random static edges to this \texttt{IntEvolvingGraph}. For example,
the second random \texttt{IntEvolvingGraph} in the sequence
has about $1.5 \times 10^8$ static edges and the third
has $1.8 \times 10^8$ static edges.
% At each time stamp $i$, a static edge is included in $G^{[i]} \in G_4$ with
% probability $p$.
Note that in this experiment, we do not have direct control
over the full edge set $E$, only the static edge set $\tilde E$. 
When we add new static edges, new causal edges may be added as well
(if the corresponding temporal nodes were not active before).
However, the number of newly introduced causal edges for each active node 
is bounded by the number of time stamps, so it suffices to demonstrate
linear scaling in $|\tilde E|$.
Figure \ref{fig:time} shows the plots of number of edges against the computation time
for running Algorithm \ref{alg:bfs} in Julia. All experiments are conducted
on a single core of a Linux system with an Intel(R) Xeon(R) E7-8850 CPU running
at 2.00 GHz clock speed.
The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with the result of Theorem \ref{thm:complexity}.

>>>>

It is well known that sparse matrix-vector product is equivalent to a
one-step BFS on the corresponding (static) graph~\cite[Sec. 1.1]{kegi11}.
In this paper, we demonstrate a correct corresponding result for an evolving
graph by constructing a block triangular matrix representation of the graph that
takes into account both static and causal connections between active nodes.
In Section~\ref{sec:breadth-first-search}, we explain how the BFS algorithm can
be applied to an evolving graph to enumerate paths that traverse edges across time
and space. Section~\ref{sec:temporal-paths} provides an example showing that considering
only products of the time-dependent adjacency matrices fails to enumerate certain
temporal paths. We present and demonstrate the BFS algorithm over evolving graphs
in Section~\ref{sec:evolving-graph-bfs}, showing its formal equivalence to
BFS over a particular static graph generated by adding causal edges that connect
active nodes. This static graph generates an algebraic
representation of the BFS as power iteration of its adjacency matrix to a starting
search node, as shown in Section~\ref{sec:bfsla}. The algebraic formulation also
demonstrates interesting connections between properties of the BFS algorithm and
the adjacency matrix. We describe in Section~\ref{sec:implementation-julia} an
implementation of the algorithm in Julia. Finally in Section~\ref{sec:applications},
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

The node set $V$ and edge set $E$ now define a static directed graph $G=(V,E)$
that is in 1-1 correspondence with the evolving graph $G_n$. The node set $V$ of $G$ is in 1-1 correspondence with active nodes of $G_n$ while the edge set $E$ is in 1-1
correspondence with all temporal paths of length 2 on $G_n$.

@@

We can then represent $G_n$ using a sequence of adjacency matrices
$A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$. The example in
Figure~\ref{fig:eg_shortest_path} can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle.
\]

@@

We generate a sequence of random (directed) \texttt{IntEvolvingGraph}s
with $10^5$ active nodes and $10$ time stamps.
The first \texttt{IntEvolvingGraph} in the sequence has about
$10^8$ static edges.
We consecutively add new
random static edges to this \texttt{IntEvolvingGraph}. For example,
the second random \texttt{IntEvolvingGraph} in the sequence
has about $1.5 \times 10^8$ static edges and the third
has $1.8 \times 10^8$ static edges.
% At each time stamp $i$, a static edge is included in $G^{[i]} \in G_4$ with
% probability $p$.
Note that in this experiment, we do not have direct control
over the full edge set $E$, only the static edge set $\tilde E$.
When we add new static edges, new causal edges may be added as well
(if the corresponding temporal nodes were not active before).
However, the number of newly introduced causal edges for each active node
is bounded by the number of time stamps, so it suffices to demonstrate
linear scaling in $|\tilde E|$.
Figure \ref{fig:time} shows the plots of number of edges against the computation time
for running Algorithm \ref{alg:bfs} in Julia. All experiments are conducted
on a single core of a Linux system with 1TB of RAM and 80 cores of Intel(R)
Xeon(R) E7-8850s running at 2.00 GHz clock speed.
The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with the result of Theorem \ref{thm:complexity}.
